---
title: "My Most Important Health Habit"
author: "Paul Brennan, Cardiff University, BrennanP@cardiff.ac.uk"
date: "28 March 2017"
output: ioslides_presentation
---

## Exploring Health Habits

Aim: to developing an interactive visualisation about the best healthy habit for me... 


## Good Healthy Habits
- Don't smoke
- Be physically active
- Enjoy a good diet - fruit & veg; refrain from red meat 
- Try to control your weight
- Don't drink too much

## Show me the money.... well the numbers...
- epidimology: study of large populations. 
- I'm going to focus on a study of physical activity.
-"Physical activity and all-cause mortality..." Ekelund et al, 2015
- Cohort study in 334,161 European men and women. 

## Hazard ratios...
- <https://en.wikipedia.org/wiki/Hazard_ratio>
- Hazard ratio > 1, this habit will reduce life expectancy. 
- Smoking 20 a day = hazard ratio of 2.2. 
- Being physically active = hazard ratio of 0.66

```{r, echo = FALSE, warning=FALSE, message=FALSE}
# smoking 20 a day - Hazard ratio 2.2, change +8
smoke20hazratio <- 2.2
# hazard ratio = h
# each incrased year of age associated with hazard ratio r 
# increased risk associated with the risk factor is equivalent to being
# a change in t years of your effective age if
# r^t = h

# t = log(h)/log(r)

changeEffAge <- function(hazRatio){log(hazRatio)/log(1.1)}
```

## Convert Hazard ratios to life years... 
- So Professor David Spiegelhalter has info that allows us to interpret this...  
- <https://understandinguncertainty.org/what-your-effective-age>
- converted hazard ratio to change in life expectancy in years. 
- Smoking 20 a day - hazard ratio of 2.2 - reduces life expectancy by `r round(changeEffAge(smoke20hazratio), 1)` years. 
- Being physically active can increase life expectancy by `r -round(changeEffAge(0.76), 1)` years.
- Easier to relate to than hazard ratios

## But depends on OTHER factors... 
- statisticians call significant interactions.
- example, there is an interaction between physical activity and body mass index. 
- Depending on your weight, the benefit from physical activity is different. 

## Core concept:
- Certain habits may be worth more 'added life' than others 
- depending on your lifestyle...
- gender...
- current habits...

## I like a bit of data...
So I have started with a graph...

```{r, echo = FALSE, warning=FALSE, message=FALSE}
library(RCurl)
library(ggplot2)
# here is a link for all the normalised data from the paper (>50000 probes) 
x <- getURL("https://raw.githubusercontent.com/brennanpincardiff/RforBiochemists/master/data/exercise_hzRatio2.csv")

# turn the data into a useful data.frame
data <- read.csv(text = x, header = TRUE)

bmi_no2 <- factor(data$bmi_no, levels = c("18.5-24.9", "25-30", ">30"))
data$bmi_no <- bmi_no2

# activity categories
actCats <- c("Moderately inactive", "Moderately active", "Active")

ggplot(data,
       aes(x = ActivityLevel,
           y = AddedYearsPos,
           fill = bmi_no)) +
  geom_bar(stat="identity", position="dodge")  +
  ylab("Added Years Possible") +
  xlab("Activity level (change relative to Inactive)") +
  facet_wrap(~gender) +
  scale_x_discrete(limits = actCats) + 
  theme_bw()

```

## So I have turned the graph into a Shiny App
- can have a look in R Studio
- have a server.R file  (downloads data; does calculations...)
- ui.R file   (shows the data, allows responses...)

## demo this and get feedback...
<https://brennanpcardiff.shinyapps.io/health_habits_app_v03/>



## Sources & Inspiration
- Elkund et al (2015) Am J Clin Nutr. 101:613-21
- Wareham et al (2002) Public Health Nutrition 6: 407–413
- Prof David Spiegelhalter
- His book: The Norm Chronicles
- His Blog: <https://understandinguncertainty.org/what-your-effective-age>

## Thanks
- Twitter: @brennanpcardiff
- <http://rforbiochemists.blogspot.co.uk/>


## BMI
- BMI was calculated as body weight (in kg) divided by height squared (in m).
- Individuals were categorized into 
- normal-weight (BMI: 18.5–24.9), overweight (BMI: 25–30), and obese (BMI >30)
