---
title: "R for Biochemists"
subtitle: "Introduction"
author: "Dr Paul Brennan"
date: "26 May 2015"
output: ioslides_presentation
---

## Welcome to R for Biochemists
- Aim: show how to use R to analyse biochemical data.
- Step 1: some basic functions
- Example: a protein assay
- [R for Biochemists Blog](http://rforbiochemists.blogspot.co.uk/)

## First step: getting some data into R
```{r}
# Use the assignment function "<-"
# Protein Concentrations
prot <- c(0.000, 0.016, 0.031, 0.063, 0.125, 0.250, 0.500, 1.000, 
          0.000, 0.016, 0.031, 0.063, 0.125, 0.250, 0.500, 1.000) 

# Absorbance from my protein assay
abs <- c(0.329, 0.352, 0.349, 0.379, 0.417, 0.491, 0.668, 0.956, 
         0.327, 0.341, 0.355, 0.383, 0.417, 0.446, 0.655, 0.905)

```

## Plot the data
```{r, echo=TRUE}
plot(abs~prot)
```

## Add a line
```{r, echo=TRUE, eval=FALSE,  results='hide'}
plot(abs~prot)
#Calculate the line using the linear model function
line <- lm(abs~prot)
#Draw the line
abline(line)
```

## Add a line
```{r, echo=FALSE, eval=TRUE,  results='asis'}
plot(abs~prot)
#Calculate the line using the linear model function
line <- lm(abs~prot)
#Draw the line
abline(line)
```

## Essential concepts
- <b>Functions</b> - these do things
- Example: plot()
- <b>Objects</b> - these are things
- Example: prot is a vector of numbers
- We links these together to write a script

## Improving our plots
- functions have <b>Arguments</b>
- example = xlab
```{r, echo=TRUE, eval=FALSE,  results='hide'}
plot(abs~prot, 
     xlab = "[Protein] (microg/ml)",
     ylab = "Absorbance (570nm)",
     main = "Protein Assay 20th April 2015")
abline(line)
```
## Improving our plots
```{r, echo=FALSE, eval=TRUE,  results='asis'}
plot(abs~prot, 
     xlab = "[Protein] (microg/ml)",
     ylab = "Absorbance (570nm)",
     main = "Protein Assay 20th April 2015")
abline(line)
```

## Adding R-squared to graph
- R does good statistics
- Extract correlation coefficient out of the line object
```{r, echo=TRUE, eval=FALSE,  results='hide'}
# Extract the correlation coefficient out of the line object
r2 <- round(summary(line)$r.squared, 3)
# Create the label
mylabel = bquote(italic(R)^2 == .(format(r2, digits = 3)))
# Put the label on the graph with the text() function
text(x = 0.2, y = 0.9, labels = mylabel)

```
## Adding R-squared to graph
```{r, echo=FALSE, eval=TRUE,  results='asis'}
plot(abs~prot, 
     xlab = "[Protein] (microg/ml)",
     ylab = "Absorbance (570nm)",
     main = "Protein Assay 20th April 2015")
abline(line)
r2 <- round(summary(line)$r.squared, 3)
mylabel = bquote(italic(R)^2 == .(format(r2, digits = 3)))
text(x = 0.2, y = 0.9, labels = mylabel)
```

## Calculate some unknowns - 1
```{r, echo=TRUE, eval=FALSE,  results='hide'}
#Equation of a line y = mx + c
#In our case abs = slope * prot + intercept
# ukn.prot = (abs - intercept)/slope
# Extract the correlation coefficient out of the line object
int <- summary(line)$coefficients[1]
slope <- summary(line)$coefficients[2]
# Create the label of the line
mylabel = bquote(y == .(format(slope, digits = 3))*x + .(format(int, digits = 3)))
# Put the label on the graph with the text() function
text(x = 0.2, y = 0.8, labels = mylabel)
```

## Calculate some unknowns - 2
```{r, echo=TRUE, eval=FALSE,  results='hide'}
#now calculate some unknown protein concs from absorbances
#put the unknowns into a vector
abs.unknowns <- c(0.554, 0.568, 0.705)
#rearrange the equation of the line to 
# ukn.prot = (abs - intercept)/slope
prot.unknowns <- (abs.unknowns - int)/slope
```

## Put the unknowns on graph
```{r, echo=FALSE, eval=TRUE,  results='asis'}
#Improve the graph:
plot(abs~prot, 
     xlab = "[Protein] (microg/ml)",
     ylab = "Absorbance (570nm)",
     main = "Protein Assay 20th April 2015")
abline(line)

r2 <- round(summary(line)$r.squared, 3)
mylabel = bquote(italic(R)^2 == .(format(r2, digits = 3)))
text(x = 0.2, y = 0.9, labels = mylabel)


#Equation of a line y = mx + c
#In our case abs = slope * prot + intercept
# ukn.prot = (abs - intercept)/slope
int <- summary(line)$coefficients[1]
slope <- summary(line)$coefficients[2]
mylabel = bquote(y == .(format(slope, digits = 3))*x + .(format(int, digits = 3)))
text(x = 0.2, y = 0.8, labels = mylabel)

#now calculate some unknown protein concs from absorbances
#put the unknowns into a vector
abs.unknowns <- c(0.554, 0.568, 0.705)
#rearrange the equation of the line to ukn.prot = (abs - intercept)/slope
prot.unknowns <- (abs.unknowns - int)/slope

#put the answers on the graph
text(x = 0.8, y = (0.6), "Abs")
text(x = 0.92, y = (0.6), "Prot")
for (i in 1:length(abs.unknowns)){
  text(x = 0.8, y = (0.6 - i/20), abs.unknowns[i])
  text(x = 0.92, y = (0.6 - i/20), round(prot.unknowns[i], 3))
}
```

## Summary
- We have put data into R - assignment function "<-"
- Done a nice plot - plot()
- Created a line object - lm()
- Extracted data out of this object - summary(line)$r.squared
- Calculated unknowns

