---
title: "Data Structures"
author: "Dr Paul Brennan, Cardiff University" 
date: "27 May 2015"
output: ioslides_presentation
---

## Welcome to R for Biochemists
- Aim: show how to use R to analyse biochemical data.
- Step 2: exploring data structures
- Example: a protein assay
- [R for Biochemists Blog](http://rforbiochemists.blogspot.co.uk/)
- http://medicine.cf.ac.uk/person/paul-brennan/

## Data goes into objects
```{r}
# Use the assignment function "<-"
# Protein Concentrations
prot <- c(0.000, 0.016, 0.031, 0.063, 0.125, 0.250, 0.500, 1.000, 
          0.000, 0.016, 0.031, 0.063, 0.125, 0.250, 0.500, 1.000) 

# Absorbance from my protein assay
abs <- c(0.329, 0.352, 0.349, 0.379, 0.417, 0.491, 0.668, 0.956, 
         0.327, 0.341, 0.355, 0.383, 0.417, 0.446, 0.655, 0.905)

```

## These objects are vectors
### They appear in the R-Studio environment as Values
```{r, echo=TRUE, eval=TRUE,  results='asis'}

str(prot)

str(abs)
```

## These objects are vectors
* A vector is the simplist type of object
* All the data elements must be the same type
* Can be numeric, character, logical, factors
```{r, echo=TRUE, eval=TRUE,  results='markup'}
class(prot)

```

## Other classes of variables
```{r, echo=TRUE, eval=TRUE,  results='asis'}
protein <- "albumin"
class(protein)

truth <- c(TRUE, FALSE, TRUE, TRUE)
class(truth)
```


## To output vectors
### write the name of the vector
```{r, echo=TRUE, eval=TRUE,  results='markup'}
prot
```

## To identify elements inside the vector
```{r, echo=TRUE, eval=TRUE,  results='asis'}
prot[2]  # second element of the vector

### or parts of objects
prot[1:8]  # elements 1 to 8 of the vector
```

## Functions can be applied to whole objects 
```{r, echo=TRUE, eval=TRUE,  results='asis'}
#plot() function uses the elements in order
plot(abs~prot)
```

# More Complicated Structures
## Lists are another type of object
* The lm() function makes an object called "line" 
* "line" is a list. 
* Lists can contain mixture of data classes. 
```{r, echo=TRUE, eval=TRUE,  results='markup'}
line <- lm(abs~prot)
### not the assignment function "<-"
### the R-Studio environment says a "List of 12"
```

## Extracting information from lists
### There are various ways 
```{r, echo=TRUE, eval=TRUE,  results='markup'}
 
## type the name of the object
line
```

## Extracting information from lists
```{r, echo=TRUE, eval=TRUE,  results='markup'}
summary(line)   ## use the summary() function
```

## Extracting information from lists
```{r, echo=TRUE, eval=TRUE,  results='asis'}
## use the $  - VERY USEFUL
summary(line)$r.squared
```


## We used $ to extract the r2 for the graph
### We created the object r2 using the function summary()
```{r, echo=TRUE, eval=TRUE,  results='asis'}
r2 <- summary(line)$r.squared
### and the function round() - gives us three decimal points
r2 <- round(summary(line)$r.squared, 3)
r2
class(r2)
### from the list we have extracted a number. 
```


## Matrices are two dimensional structures 
###  Data classes are all the same

## Data frames are two dimensional structures
###  Contains different classes of data


## When we import data, it often becomes a data frame.
* here is an example:
```{r, echo=TRUE, eval=TRUE,  results='asis'}
data <- read.csv("http://science2therapy.com/data/wellsDataSimp.csv")
```
### the R-Studio environment puts it in "data" and gives us some info

## Have a look at it
```{r, echo=TRUE, eval=TRUE,  results='markup'}
data
```

## Look at the structure 
* Names of columns 
* Class of the data within the column

```{r, echo=TRUE, eval=TRUE,  results='markup'}
str(data)
```
### note: Factors, num, int

## Extracting information from data frames
* use $ and the column name
```{r, echo=TRUE, eval=TRUE,  results='markup'}
data$Virus
```

## Simple plot from this data frame
```{r, echo=TRUE, eval=TRUE,  results='asis'}
plot(data[5:7])  # these are the column numbers
```

## Another plot from this data frame
```{r, echo=TRUE, eval=TRUE,  results='asis'}
plot(data$P.Erk, data$S.phase.cnt)
```

## Summary
- Vectors - simple object, data same class
- Lists - one dimensional object, data different class
- Extract data out of List - summary(line)$r.squared
- Data frames - two dimensional object, data different class
- Extract data out of data
```{r, echo=TRUE, eval=FALSE,  results='asis'}
plot(data$P.Erk, data$S.phase.cnt)
```

## Next tutorial
### Manipulating objects including data frames 
